---
title: "Data transformation with dplyr"
output: html_notebook
---

```{r}
library(tidyverse)
library(nycflights13)
```

## Filter

### Exercise 1-a
Flights with arrival delay of 2hrs or more.
```{r}
flights %>% select(flight, arr_delay) %>%
  filter(arr_delay >= 120) %>%
  arrange(desc(flight))
# 10200
```

### Exercise 1-b
Flights to Houston (IAH or HOU)
```{r}
# glimpse(flights)

flights %>% select(flight, dest) %>% 
  filter(dest %in% c("IAH", "HOU")) %>% 
  arrange(desc(flight))
# 9313
```

### Exercise 1-c
Flights operated by United, American or Delta
```{r}
# Which flight operators are present?
flights %>% distinct(carrier) %>% arrange(desc(carrier))

# Filter UA, AA, DL
flights %>% filter(carrier %in% c("UA", "AA", "DL"))
# 139504 flights
```

### Exercise 1-d
Flights departed in summer (July, August, September)
```{r}
flights %>% filter(month %in% c(7, 8, 9))
# 86326 flights

# alternative methods
# flights %>% filter(month >= 7 & month <= 9)
# flights %>% filter(between(month, 7, 9))
```


### Exercise 1-e
Arrived more than 2 hours later but didn't leave late.
```{r}
flights %>% filter(arr_delay > 120 & dep_delay == 0)
# 3 flights
```

### Exercise 1-f
Delayed by at least an hour but made up at least 30 minutes in flight.
```{r}
flights %>% filter(dep_delay > 60 & arr_delay < -30)
# No flights
```

### Exercise 1-g
Departed between midnight and 6am, inclusive.
```{r}
# Visualise flight distribution
# ggplot(flights, aes(x = dep_time)) +
#  geom_histogram(binwidth = 60)

flights %>% filter(dep_time >= 0 & dep_time <= 600)
# 9344 flights
```


### Exercise 2

See 1-d above

### Exercise 3

How many flights have missing `dep_time`? 

```{r}
flights %>% filter(is.na(dep_time))
# 8255 flights with NA dep_time. These flights also have other missing variables such as dep_delay that are indicative of the flight actually taking place. These flights were likely cancelled.
```


### Exercise 4

```{r}
NA ^ 0
# 1

NA | TRUE
# TRUE

FALSE & NA
# FALSE

NA * 0
# NA
```


## Arrange

### Exercise 1
Sort NAs to start.
```{r}
flights %>% arrange(desc(is.na(dep_delay)))
```

### Exercise 2
Most delayed flights.
```{r}
flights %>% arrange(desc(dep_delay))
```

Flights that left earliest.
```{r}
flights %>% arrange(desc(desc(dep_delay)))
```


### Exercise 3
Sort to find fastest flights.
```{r}
flights %>% arrange(desc(distance/air_time))
```


### Exercise 4
Longest travelled flights.
```{r}
flights %>% arrange(desc(distance))
# JFK to HNL the longest
```

Shortest travelled flights.
```{r}
flights %>% arrange(desc(desc(distance)))
# EWL to PHL the shortest of flights that weren't cancelled
```

## Select

### Exercise 1
Different select routes for dep_time, dep_delay, arr_time, and arr_delay from flights.
```{r}
# glimpse(flights)

library(microbenchmark)
dplyr1 <- flights %>% select(dep_time, dep_delay, arr_time, arr_delay)
dplyr1_2 <- select(flights, dep_time, dep_delay, arr_time, arr_delay)
dplyr2 <- flights %>% select(starts_with("dep_"),  starts_with("arr_"))
dplyr2_2 <- select(flights, starts_with("dep_"),  starts_with("arr_"))

# Base R
vars <- c("dep_time", "dep_delay", "arr_time", "arr_delay")
base1 <- flights[vars]

# Benchmark 
bench <- microbenchmark(dplyr1, dplyr1_2, dplyr2, dplyr2_2, base1, times = 10000L)
bench
rm(list = ls())
```

Of the different approaches dplyr select without piping using helper functions is slightly faster on average - but only by a few nanoseconds.

### Exercise 2
Include variable multiple times
```{r}
flights %>% select(dep_delay, dep_delay)
```
Only return one column.

### Exercise 3
`one_of()` selects from a character vector. This vector could be passed directly to `select`.
```{r}
vars <- c("year", "month", "day", "dep_delay", "arr_delay")
flights %>% select(one_of(vars))

```


### Exercise 4
The following is not case sensitive.
```{r}
select(flights, contains("TIME"))
```

The select helper `contains` can be ammended to be case sensitive.
```{r}
select(flights, contains("TIME", ignore.case = FALSE))
```
But there are no columns with TIME in their name.


## Mutate and transmute

### Exercise 1
Convert `dep_time` and `sched_dep_time` (presented as hhmm) into total number of minutes.
```{r}
# glimpse(flights)

# integer division (%/%) of hhmm format to get number of hours then multiply by 60 for minutes
# then add remainder (%%) minutes
flights %>% transmute(dep_time_mins = dep_time %/% 100 * 60 + dep_time %% 100,
                      sched_dep_time_mins = sched_dep_time %/% 100 * 60 + sched_dep_time %% 100)
```

### Exercise 2
Compare `air_time` with `arr_time`-`dep_time`. 

They should be the same. Let's check...
```{r}
flights %>% transmute(air_time, arr_dep_time = arr_time - dep_time)
```
They are not the same. 

Why?

Arrival and depart times are in hhmm format so we can convert these into pure minutes.
```{r}
# glimpse(flights)

# Function to convert hhmm to mins
hhmm2mins <- function(x) {
  xhrs <- x %/% 100 * 60
  xmm <- x %% 100
  xmins <- xhrs + xmm
  xmins
}

flights %>% transmute(carrier, flight, air_time, 
                      arr_time, dep_time,
                      arr_dep_mins = hhmm2mins(arr_time) - hhmm2mins(dep_time)) %>%
  mutate(air_time_diff = arr_dep_mins - air_time) %>% 
  select(-arr_time, -dep_time) %>% 
  arrange(desc(air_time_diff))
```
There is still a difference.
Are there any instances where two methods are the same?

```{r}
# glimpse(flights)

flights %>% transmute(carrier, flight, origin, dest, air_time, 
                      arr_time, dep_time,
                      arr_dep_mins = hhmm2mins(arr_time) - hhmm2mins(dep_time)) %>%
  mutate(air_time_diff = arr_dep_mins - air_time) %>% 
  select(-carrier, -flight) 
# %>% filter(air_time == arr_dep_mins)
```
I think there are two problems:
1. Some flights are not within the same time zone - RITA, Bureau of transportation statistics, http://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=236 says all times are given in local time zone.
2. There are taxi times on the runway to be taken into account - not included in `air_time` variable.
3. Some flights occur overnight, this needs to be taken into account when calculating total flight times.

### Exercise 3
Comparing `dep_time`, `sched_dep_time` and `dep_delay`. `dep_delay` should be `dep_time` - `sched_dep_time`.
```{r}
flights %>% select(flight, origin, dest, dep_time, sched_dep_time, dep_delay) %>% 
  mutate(dep_delay_calc = hhmm2mins(dep_time) - hhmm2mins(sched_dep_time)) %>% 
  arrange(dep_delay_calc)
```
However, overnight flights create a problem with performing simple subtraction of `dep_time` - `sched_dep_time` to obtain flight delays for all journeys.

### Exercise 4
Find the 10 most delayed flights using ranking.

```{r}
delayed <- flights %>% transmute(flight, carrier, dep_delay, arr_delay, rank = min_rank(desc(arr_delay))) %>% arrange(rank)


delayed %>% mutate(hhmm2mins(dep_delay))

ggplot(delayed, aes(x = dep_delay )) + 
  geom_histogram(bins = 200)
  
View(flights)
  
```






















